This version of the \textit{CHERI Instruction-Set Architecture} is a full
release of the Version 7 specification:

\begin{itemize}
\item We have now deprecated the CHERI-128 capability compression format, in
  favor of CHERI Concentrate.

\item The RISC-V \insnnoref{AUIPC} instruction now returns a
  \PCC{}-relative capability in the capability encoding mode.

\item Capabilities now contain a \cflags{} field (\cref{sec:model-flags}),
  which will hold state that
  can be changed without affecting privilege.
  Corresponding experimental \insnref{CGetFlags} and
  \insnref{CSetFlags} instructions have been added.

\item The capability encoding-mode bit in CHERI-RISC-V is specified as a bit
  in the \cflags{} field of a capability.
  The current mode is defined as the flag bit in the currently installed
  \PCC{}.
  Design considerations and other potential options are described in
  Chapter~\ref{chap:rationale}.

\item We now more explicitly describe the reset states of special- and
  general-purpose capability registers for CHERI-MIPS and CHERI-RISC-V.

\item Compressed capabilities now contain a dedicated \cotype{} field that
  always holds an object type (see
  \cref{sec:model-object-types,section:object-type}), rather than stealing
  bounds bits for object type when sealing.  Now, any representable capability
  may be sealed.  Several object type values are reserved for architectural
  experimentation (see \cref{tab:archotypes}).

\item More detail is provided regarding the integration of CHERI Concentrate
  with special registers, its alignment requirements, and so on.

\item Initial discussion of a disjoint capability tree for physical
  addresses and hardware facilities using these has been added to
  the experimental appendix, in \cref{app:exp:physcap}.

\item Initial discussion of a hybrid 64/128-bit capability design has been
  added to the experimental appendix, in \cref{sec:windowedshortcaps}.

\item We have added formal Sail instruction semantics for CHERI-RISC-V; this
  is currently in Appendix~\ref{app:isaquick-riscv}.

\item We have added a reference to our IEEE TC 2019 paper, \textit{CHERI
  Concentrate: Practical Compressed Capabilities}, which describes our current
  approach to capability compression.

\item We have added a reference to Alexandre Joannou's PhD dissertation,
  \textit{High-perform\-ance memory safety: optimizing the CHERI capability
  machine}, which describes approaches to improving the efficiency of
  capability compression and tagged memory.

\end{itemize}
