\chapter{The CHERI-MIPS Instruction-Set Reference}
\label{chap:isaref-mips}

CHERI-MIPS's instructions express a variety of operations affecting capability
and integer registers as well as memory access and control flow.
A key design concern is {\em guarded manipulation}, which provides
a set of constraints
across all instructions that ensure monotonic non-increase in rights through
capability manipulations.
These instructions also assume, and specify, the presence of {\em tagged
memory}, described in the previous chapter, which protects in-memory
representations of capability values.
Many further behaviors, such as reset state and exception handling (taken for
granted in these instruction descriptions), are also described in the previous
chapter.
A small number of more recently specified experimental instructions are
specified in Appendix~\ref{app:experimental} rather than in this chapter.

The instructions fall into a number of categories: instructions to copy
fields from capability registers into integer registers so that they
can be computed on, instructions for refining fields within capabilities, instructions for memory access via
capabilities, instructions for jumps via capabilities, instructions for sealing capabilities, and instructions for capability invocation.
In this chapter, we specify each instruction via both informal descriptions
and code in the Sail language.
To allow for more succinct code descriptions, we rely on a number of
common function definitions also described in this chapter.

%TODO: Insert this section
%\section{Capability Exceptions}

\newcommand{\cchecktag}{\emph{cb}.\ctag{} is not set.}

\section{Sail language used in instruction descriptions}
\label{sec:sail-language-description}

\newcommand{\isail}[1]{\lstinline[language=sail]{#1}}

The instruction descriptions contained in this chapter are accompanied
by code in the Sail language~\cite{sail-popl2019,sail-url} taken from the Sail CHERI-MIPS implementation~\cite{sail-cheri-mips}.  Sail is
a domain specific imperative language designed for describing
processor architectures.  It has a compiler that can output executable
code in OCaml or C for building executable models, and can also
translate to various theorem prover languages for automated reasoning
about the ISA.

The following is a brief description of the Sail language features
used in this document. For a full description see the Sail
language documentation.

Types used in Sail:

\label{sailMIPSzbits}
\label{sailRISCVzbits}
\begin{itemize}
\item \isail{int} Sail integers are of arbitrary precision (therefore there are no overflows) but can be constrained using simple first-order constraints. As a common case integer range types can be defined using \isail{range(a,b)} to indicate an integer in the range $a$ to $b$ inclusive. Operations on integers repsect the constraints on their operands so, for example, if \isail{x} and \isail{y} have type \isail{range(a, b)} then \isail{x + y} has type \isail{range(a + a, b + b)}. Integer literals are written in decimal.
\item \isail{bits(n)} \label{zbits} is a bit vector of length \isail{n}. Vectors are indexed using square bracket notation with index 0 being the least significant bit. Aritmetic and logical operations on vectors are defined on two vectors of equal length producing a result of the same length and truncating on overflow. Where signedness is significant it is indicated in the operator name, for example \isail{<_s} performs signed comparison of bit vectors . Bit vector literals are written in hexadecimal for multiples of four bits or in binary with \isail{0x} or \isail{0b} prefixes, e.g. \isail{0x3} means `0011' and \isail{0b11} means `11'. The at symbol, \isail{@}, indicates concatenation of vectors.
\item \isail{structs} are similar to C structs with named, typed fields acessed with a dot as in \isail{struct_val.field_name}. Struct copying with field updates is also supported as in \isail{\{struct_val with field_name=new_val\}}.
\item Registers in Sail contain the architectural state that is modified by instruction execution. By convention register names in the CHERI specification start with a capital letter to distinguish them from local variables. Sail also supports a form of `assignment' to function calls as in \isail{wGPR(rd) = result}. This is just syntactic sugar for an extra argument to the function call. This syntax is used by functions that write registers or memory and have special behavior such as \isail{wGPR}, \isail{writeCapReg} and \isail{MEMw}.
\end{itemize}

The following operators and expression syntax are used in the Sail code:

\begin{itemize}
\item \label{sailRISCVznot}\label{sailMIPSznot}Boolean operators:
\isail{not}, \isail{|} (logical OR), \isail{&} (logical AND), \isail{^} (exclusive OR)

\item Integer operators:
\isail{+} (addition), \isail{-} (subtraction), \isail{*} (multiplication), \isail{\%} (modulo)

Sail operations on integers are the usual mathematical operators. Note \isail{a \% b} is the modulo operator that, for $b > 0$ returns a value in the range $0$ to $b-1$ regardless of the sign of $a$. Although Sail integers are notionally infinite in range, CHERI instructions can be implemented with finite arithmetic.

\item Bit vector operators:
\isail{&} (bitwise AND), \isail{<_s} (signed less than), \isail{@} (bit vector concatenation)

\item Equality:
\isail{==} (equal), \isail{!=} (not equal)

\item Vector slice:

  \isail{v[a..b]}

Creates a sub-range of a vector from index $a$ down to $b$ inclusive.

\item Local variables:

  \isail{mutable_var = exp;} \\
  \isail{let immutable_var = exp;}

Mutable variables are introduced by simply assigning to them. An explicit type may be given following a colon, but types can usually be inferred. Sail supports mutable or immutable variables where immutable ones are introduced by \isail{let} and assigned only once when created.

\item Functional if:

\isail{if cond then exp1 else exp2}

May return a value similar to C ternary operator.


\item Foreach loop:

% XXX: for som reason sail generates a link to "to" for ccleartags/cloadtags
\begin{lstlisting}[language=sail,label=sailMIPSzto]
foreach(i from start_exp to end_exp) {
  body
};
\end{lstlisting}

\item Function invocation:

\isail{func_id (arg1, arg2)}

\item Field selection from struct:

  \isail{struct\_val.field}

Returns the value of the given field from structure.

\item Functional update of structure:

\isail{\{struct\_val with field=exp\}}

A copy of the structure with the named field replaced with another value.

\end{itemize}

% rmn30: removing these sections for now. Might be nice to update them for Sail.
%The precedence of the operators used in the pseudocode is shown in
%Table~\ref{table:operator-precedence}.
%
%\begin{figure}
%\begin{center}
%\begin{tabular}{l}
%\algorithmicor{} \\
%\algorithmicand{} \\
%\algorithmicnot{} \\
%$=$, $\ne$, $<$, $\le$, $>$, $\ge$ \\
%\algorithmicwith{} \\
%$+$, $-$ \\
%$*$, $/$, $\bmod$ \\
%$a^{b}$
%\end{tabular}
%\end{center}
%\caption{Operator precedence in pseudocode}
%\label{table:operator-precedence}
%\end{figure}

\section{Common Constant Definitions}

The constants used in the Sail are show in \cref{table:pseudocode-constants}.
% ; their value depends on the capability format in use and architecture specific features such as the RISC-V capability mode flag.

\begin{figure}
\begin{center}
\begin{tabular}{lrl}
\toprule
 & 128-bit & Description \\
\midrule
\verb+cap_size+ & 16 & Number of bytes used to store a capability. \\
%\verb+caps_per_cacheline+ & - & Number of capabilities in a cacheline (uArch detail) \\
\verb+max_otype+ & $2^{18}-1$ & Maximum \cotype{} allowed by capability format. \\
\verb+last_hperm+ & 11 & Last hardware-defined permission bit. \\
\verb+first_uperm+ & 15 & First software-defined permission bit. \\
\verb+last_uperm+ & 18 & Last software-defined permission bit. \\
\verb+num_flags+ & 0 & Number of capability flags. \\
\bottomrule
\end{tabular}
\end{center}

\caption{Constants in Sail code}
\label{table:pseudocode-constants}
\end{figure}

%The null capability is defined as follows:
%
%null\_capability = int\_to\_cap(0)
%
%\textbf{TO DO}: We should have a table defining the values of the capability
%exception codes.

%\section{Common Variable Definitions}
%
%The following variables are used in the pseudocode:
%
%\begin{tabular}{lll}
%cb & : & Capability \\
%cd & : & Capability \\
%cs & : & Capability \\
%ct & : & Capability \\
%rd & : & Unsigned64 \\
%rs & : & Unsigned64 \\
%rt & : & Unsigned64 \\
%mask & : & Unsigned16 \\
%offset & : & Signed16 \\
%\end{tabular}
%
%\textbf{TO DO: The notation for types should be explained as well. \emph{mem}
%and \emph{tags} are also used as variables.}

\section{Common Function Definitions}

This section contains descriptions of convenience functions used by the Sail code featured in this chapter.

\subsection*{Functions for integer and bit vector manipulation}
\label{sec:sail-int-and-bitvector-functions}

The following functions convert between bit vectors and integers and manipulate bit vectors:

\medskip
\sailMIPSval{unsigned}
\sailMIPSval{signed}
\sailMIPSval{to\_bits}
\sailMIPSval{truncate}
\sailMIPSval{pow2}

% The following are overloads so we can't easily use the generated latex
% Hacky hspace to get rid of unwanted hangindent

\phantomsection
\label{sailMIPSzzzerozyextend}
\saildocval{Adds zeros in most significant bits of vector to obtain a vector of desired length.}{\hspace{-\parindent}\isail{zero_extend}}

\label{sailMIPSzsignzyextend}
\saildocval{Extends the most significant bits of vector preserving the sign bit.}{\hspace{-\parindent}\isail{sign_extend}}

\label{sailMIPSzzzeros}
\saildocval{Produces a bit vector of all zeros}{\hspace{-\parindent}\isail{zeros}}

\label{sailMIPSzzones}
\saildocval{Produces a bit vector of all ones}{\hspace{-\parindent}\isail{ones}}

\subsection*{Functions for ISA exception behavior}

\sailMIPSval{SignalException}
\sailMIPSval{SignalExceptionBadAddr}
\sailMIPSval{raise\_c2\_exception}
\sailMIPSval{raise\_c2\_exception\_noreg}
\sailMIPSval{raise\_c2\_exception\_badaddr}
\sailMIPSval{checkCP2usable}
\sailMIPSval{pcc\_access\_system\_regs}

\subsection*{Functions for control flow}

\sailMIPSvalexecuteBranch{}
\sailMIPSvalexecuteBranchPcc{}
\sailMIPSvalsetNextPcc{}
\sailMIPSvalexecute{}

\subsection*{Functions for reading and writing register and memory}

\sailMIPSval{rGPR}
\sailMIPSval{wGPR}
\sailMIPSval{readCapReg}
\sailMIPSval{readCapRegDDC}
\sailMIPSval{writeCapReg}
\sailMIPSval{memBitsToCapability}
\sailMIPSval{capToMemBits}
\sailMIPSval{MEMr\_wrapper}\note{rmn30}{TODO rename horrible memory functions}
\sailMIPSval{MEMr\_reserve\_wrapper}
\sailMIPSval{MEMr\_tagged}
\sailMIPSval{MEMr\_tagged\_reserve}
\sailMIPSval{MEMr\_tag}
\sailMIPSval{MEMw\_wrapper}
\sailMIPSval{MEMw\_conditional\_wrapper}
\sailMIPSval{MEMw\_tagged}
\sailMIPSval{MEMw\_tagged\_conditional}
\sailMIPSval{TLBTranslate}
\sailMIPSval{TLBTranslateC}
\sailMIPSval{wordWidthBytes}
\sailMIPSval{isAddressAligned}
\sailMIPSval{extendLoad}
\sailMIPSval{getAccessLevel}
\sailMIPSval{grantsAccess}

\subsection*{Functions for manipulating capabilities}

The Sail code abstracts the capability representation using the following functions for getting and setting fields in the capability.
% XXXRW: OK change?
%  The functions have different implementations for 256-bit and 128-bit capability formats.

The base of the capability is the address of the first byte of memory to which it grants access and the top is one greater than the last byte, so the set of dereferenceable addresses is:
\[
\{ a \in \mathbb{N} \mid \mathit{base} \leq a < \mathit{top}\}
\]
Note that for 128-bit capabilities $\mathit{top}$ can be up to $2^{64}$, meaning the entire 64-bit address space can be addressed.
% XXXRW: OK change?
%, but that the 256-bit capability format has a maximum length of $2^{64} - 1$ so the last byte of the address space is inaccessible.  Capability length is defined by the relationship $\mathit{base} + \mathit{length} = \mathit{top}$.
\note{rmn30}{XXX For CHERI 256 $top$ is actually derived using $base + length$ but for 128-bit we derive $length$ using $top - base$.  For valid (tagged) capabilities we have (we hope) the invariant $0 \leq base \leq top \leq 2^{64}$, so the length is always positive and in the range $0 .. 2^{64}$ but for untagged 256-capabilities we may have $base + length \geq 2^{64}$ so we have to be careful about overflow.  For most instructions this is OK because we check the tag before using getCapTop but this is not true of CBuildCap and CTestSubset so sail code may not have correct semantics there -- need to test. For 128-bit more thought is required... }

\medskip
\sailMIPSval{getCapBase}
\sailMIPSval{getCapTop}
\sailMIPSval{getCapLength}

\noindent The capability's address (also known as cursor) and offset (relative to base) are related by:
\[
\mathit{base} + \mathit{offset}\ \mathbf{mod}\ 2^{64} = \mathit{cursor}
\]
The following functions return the cursor and offset of a capability respectively:
\note{rmn30}{Re-name cursor to address here?}

\medskip
\sailMIPSval{getCapCursor}
\sailMIPSval{getCapOffset}
\note{rmn30}{explain what happens when offset is negative? In fact it is computed modulo $2^{64}$ and always converted straight to a 64-bit vector so not important. Should maybe just return vector.}

\noindent The following functions adjust the bounds and offset of capabilities.  When using compressed capabilities not all combinations of bounds and offset are representable, so these functions return a boolean value indicating whether the requested operation was successful.  Even in the case of failure a capability is still returned, although it may not preserve the bounds of the original capability. \note{rmn30}{Need more detail here?}

\medskip
\sailMIPSval{setCapBounds}
\sailMIPSval{setCapAddr}
\sailMIPSval{setCapOffset}
\sailMIPSval{incCapOffset}

\medskip
\sailMIPSval{getRepresentableAlignmentMask}
\sailMIPSval{getRepresentableLength}

\medskip
\sailMIPSval{hasReservedOType}
\sailMIPSval{isSentryCap}
\sailMIPSval{sealCap}
\sailMIPSval{unsealCap}
\sailMIPSval{unrepCap}

\mrnote{We should explain the is\_representable function. What properties is
it required to have? All offsets between 0 and length (inclusive) must
be representable for C semantics. Is a length of zero (for any base) required
to be representable? There is also the fast representability check -- there
are in fact several different representability functions depending on where
its called.}

Capability permissions are accessed using the following functions:

\medskip
\sailMIPSval{getCapPerms}
\sailMIPSval{setCapPerms}
\note{rmn30}{could separate user and hw permissions as per existing pseudocode}
\sailMIPSval{getCapFlags}
\sailMIPSval{setCapFlags}

%\sailMIPSval{CapStruct}
%\sailMIPSval{wordWidthBytes}
%\sailMIPSval{getAccessLevel}
%\sailMIPSval{grantsAccess}
%\sailMIPSval{nullcap}
%\sailMIPSval{uintsixfour}
%\sailMIPSval{CapLen}
%\note{rmn30}{CapLen should really be $0 .. 2^{64}$ but that fails to compile -- will need to tweak sail code a bit.}

\section{Table of CHERI-MIPS Instructions}

\begin{table}
\begin{center}
\begin{tabular}{p{1.3in}p{4.6in}}
\toprule
Mnemonic & Description \\
\midrule
  CGetAddr & \insnmipsref*[cgetaddr]{Move capability address to an integer register} \\
  CGetAndAddr & \insnmipsref*[cgetandaddr]{Move capability address to an integer register, with mask} \textit{(experimental)} \\
  CGetBase & \insnmipsref*[cgetbase]{Move base to an integer register}
    \\
  CGetFlags & \insnmipsref*[cgetflags]{Move flags to an integer register} \\
  CGetLen & \insnmipsref*[cgetlen]{Move length to an integer register}
    \\
  CGetOffset & \insnmipsref*[cgetoffset]{Move offset to an integer register} \\
  CGetPerm & \insnmipsref*[cgetperm]{Move permissions to an integer
    register} \\
  CGetSealed & \insnmipsref*[cgetsealed]{Test if a capability is sealed} \\
  CGetTag & \insnmipsref*[cgettag]{Move tag bit to an integer register} \\
  CGetType & \insnmipsref*[cgettype]{Move object type to an integer
    register} \\
  CPtrCmp & \insnmipsref*[cptrcmp]{Capability pointer compare} \\
  CToPtr & \insnmipsref*[ctoptr]{Capability to integer pointer} \\
\midrule
  CAndAddr & \insnmipsref*[candaddr]{Mask address of capability}
    \textit{(experimental)} \\
  CAndPerm & \insnmipsref*[candperm]{Restrict permissions} \\
  CBuildCap & \insnmipsref*[cbuildcap]{Import a capability}
   \textit{(experimental)} \\
  CClearRegs & \insnmipsref*[cclearregs]{Clear multiple registers} \\
  CClearTag & \insnmipsref*[ccleartag]{Clear the tag bit} \\
  CCopyType & \insnmipsref*[ccopytype]{Import a capability's \cotype{}}
   \textit{(experimental)} \\
  CFromPtr & \insnmipsref*[cfromptr]{Create capability from pointer} \\
  CGetPCC & \insnmipsref*[cgetpcc]{Move PCC to capability register} \\
  CGetPCCSetOffset & \insnmipsref*[cgetpccsetoffset]{Move PCC to capability
    register with new offset} \\
  CGetPCCIncOffset & \insnmipsref*[cgetpccincoffset]{Move PCC to capability
    register and increment offset} \\
  CGetPCCSetAddr & \insnmipsref*[cgetpccsetaddr]{Move PCC to capability
    register with new address} \\
  CIncOffset & \insnmipsref*[cincoffset]{Increment offset} \\
  CIncOffsetImm & \insnmipsref*[cincoffsetimm]{Increment Offset by
    Immediate} \\
  CMove & \insnmipsref*[cmove]{Move capability} \\
  CMOVN & \insnmipsref*[cmovn]{Conditionally move capability on non-zero} \\
  CMOVZ & \insnmipsref*[cmovz]{Conditionally move capability on zero} \\
  CReadHwr & \insnmipsref*[creadhwr]{Read a special-purpose capability register} \\
  CSetAddr & \insnmipsref*[csetaddr]{Set capability address to value from register} \\
  CSetBounds & \insnmipsref*[csetbounds]{Set bounds} \\
  CSetBoundsExact & \insnmipsref*[csetboundsexact]{Set bounds exactly} \\
  CSetBoundsImm & \insnmipsref*[csetboundsimm]{Set bounds (immediate)} \\
  CSetFlags & \insnmipsref*[csetflags]{Set flags}
    \textit{(experimental)} \\
  CSetOffset & \insnmipsref*[csetoffset]{Set cursor to an offset from base} \\
  CSub & \insnmipsref*[csub]{Subtract capabilities} \\
  CWriteHwr & \insnmipsref*[cwritehwr]{Write a special-purpose capability register} \\
\bottomrule
\end{tabular}
\end{center}
\caption{Capability coprocessor instruction summary}
\label{table:capability-instruction-summary}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{p{1.3in}p{4.6in}}
\toprule
Mnemonic & Description \\
\midrule
  CL[BHWD][U] & \insnmipsref*[clbhwd]{Load integer via capability} \\
  CLC & \insnmipsref*[clc]{Load capability via capability} \\
  CLCBI & \insnmipsref*[clcbi]{Load capability via capability (big
    immediate)} \\
  CLL[BHWD][U] & \insnmipsref*[cllbhwd]{Load linked integer via capability} \\
  CLLC & \insnmipsref*[cllc]{Load linked capability via capability} \\
  CSC & \insnmipsref*[csc]{Store capability via capability} \\
  CS[BHWD] & \insnmipsref*[csbhwd]{Store integer via capability} \\
  CSC[BHWD] & \insnmipsref*[cscbhwd]{Store conditional integer via capability} \\
  CSCC & \insnmipsref*[cscc]{Store conditional capability via capability} \\
\midrule
  CBEZ & \insnmipsref*[cbez]{Branch if capability is NULL} \\
  CBNZ & \insnmipsref*[cbnz]{Branch if capability is not NULL} \\
  CBTS & \insnmipsref*[cbts]{Branch if capability tag is set} \\
  CBTU & \insnmipsref*[cbtu]{Branch if capability tag is unset} \\
  CJALR & \insnmipsref*[cjalr]{Jump and link capability register} \\
  CJR  & \insnmipsref*[cjr]{Jump capability register} \\
\midrule
  CCheckTag & \insnmipsref*[cchecktag]{Raise exception if capability tag is
    unset} \\
\midrule
  CSeal & \insnmipsref*[cseal]{Seal a capability} \\
  CCSeal & \insnmipsref*[ccseal]{Conditionally seal a capability}
   \textit{(experimental)} \\
  CUnseal & \insnmipsref*[cunseal]{Unseal a sealed capability} \\
\midrule
  CInvoke & \insnmipsref*[cinvoke]{Call into another security domain} \\
\midrule
  CGetCause & \insnmipsref*[cgetcause]{Move the capability exception cause
    register to an integer register} \\
  CGetCID & \insnmipsref*[cgetcid]{Move the architectural Compartment ID
    (CID) to an integer register} \\
  CSetCause & \insnmipsref*[csetcause]{Set the capability exception cause
    register} \\
  CSetCID & \insnmipsref*[csetcid]{Set the architectural Compartment ID
    (CID)} \\
\bottomrule
\end{tabular}
\end{center}
\caption{Capability coprocessor instruction summary, continued}
\label{table:capability-instruction-summary-2}
\end{table}

Tables~\ref{table:capability-instruction-summary}
and~\ref{table:capability-instruction-summary-2} list available capability
coprocessor instructions.

% rmn30: personally i think this section is pointless.
%\section{Details of Individual Instructions}
%
%The following sections provide a detailed description of each CHERI ISA
%instructions.
%Each instruction description includes the following information:
%
%\begin{itemize}
%\item Instruction opcode format number
%\item Assembly language syntax
%\item Bitwise figure of the instruction layout
%\item Text description of the instruction
%\item Sail code for the instruction
%\item Enumeration of any exceptions that the instruction can trigger
%\end{itemize}

\input{insn-mips/candperm}
\input{insn-mips/cbez}
\input{insn-mips/cbts}
%\input{insn-mips/cbuildcap}
\input{insn-mips/cchecktag}
\input{insn-mips/cclearregs}
\input{insn-mips/ccleartag}
%\input{insn-mips/ccopytype}
%\input{insn-mips/ccseal}
\input{insn-mips/cfromptr}
\input{insn-mips/cgetaddr}
\input{insn-mips/cgetbase}
\input{insn-mips/cgetcid}
\input{insn-mips/cgetcause}
\input{insn-mips/cgetflags}
\input{insn-mips/cgetlen}
\input{insn-mips/cgetoffset}
\input{insn-mips/cgetpcc}
\input{insn-mips/cgetpccsetoffset}
\input{insn-mips/cgetpccincoffset}
\input{insn-mips/cgetpccsetaddr}
\input{insn-mips/cgetperm}
\input{insn-mips/cgetsealed}
\input{insn-mips/cgettag}
\input{insn-mips/cgettype}
\input{insn-mips/cincoffset}
\input{insn-mips/cincoffsetimm}
\input{insn-mips/cinvoke}
\input{insn-mips/cjalr}
\input{insn-mips/cl}
\input{insn-mips/clc}
\input{insn-mips/clcbi}
\input{insn-mips/cllbhwd}
\input{insn-mips/cllc}
\input{insn-mips/cloadtags}
\input{insn-mips/cmove}
\input{insn-mips/cmovn}
\input{insn-mips/cptrcmp}
\input{insn-mips/cram}
\input{insn-mips/creadhwr}
\input{insn-mips/crrl}
\input{insn-mips/cs}
\input{insn-mips/csc}
\input{insn-mips/cscbhwd}
\input{insn-mips/cscc}
\input{insn-mips/cseal}
\input{insn-mips/csealentry}
\input{insn-mips/csetaddr}
\input{insn-mips/csetbounds}
\input{insn-mips/csetboundsexact}
\input{insn-mips/csetboundsimm}
\input{insn-mips/csetcause}
\input{insn-mips/csetcid}
\input{insn-mips/csetflags}
\input{insn-mips/csetoffset}
\input{insn-mips/csub}
\input{insn-mips/ctoptr}
\input{insn-mips/cunseal}
\input{insn-mips/cwritehwr}

\clearpage
\section{Assembler Pseudo-Instructions}

For convenience, several pseudo-instructions are accepted by the assembler.
These expand to either single instructions or short sequences of instructions.

\subsection{CGetDefault, CSetDefault}
\insnmipslabel{cgetdefault}
\insnmipslabel{csetdefault}
\subsubsection*{Get/Set Default Capability}

\asm{CGetDefault} and \asm{CSetDefault} get and set the capability register that
is implicitly employed by the legacy MIPS load and store instructions.
In the current version of the ISA, this register is special-purpose capability
register 0.

In previous versions of the ISA, \DDC{} was register \creg{0} in the main
capability register file. In these versions of the architecture, using
\insnnoref{CSetDefault} rather than an capability operation with destination
\creg{0} allowd the Clang/LLVM compiler to know that the semantics of subsequent
MIPS loads and stores will be affected by the change to \DDC{}.

\begin{asmcode}
# The following are equivalent:
  CGetDDC $c1
  CGetDefault $c1
  CReadHWR $c1, $0
\end{asmcode}

\begin{asmcode}
# The following are equivalent:
  CSetDDC $c1
  CSetDefault $c1
  CWriteHWR $c1, $0
\end{asmcode}

\subsection{CGetEPCC, CSetEPCC}
\insnmipslabel{cgetepcc}
\insnmipslabel{csetepcc}
\subsubsection*{Get/Set Exception Program Counter Capability}

Pseudo-operations are provided for getting and setting \EPCC{}. In the current
ISA, EPCC is special-purpose capability register 31.

\begin{asmcode}
# The following are equivalent:
  CGetEPCC $c1
  CReadHWR $c1, $31
\end{asmcode}

\begin{asmcode}
# The following are equivalent:
  CSetEPCC $c1
  CWriteHWR $c1, $31
\end{asmcode}

\subsection{CGetKDC, CSetKDC}
\insnmipslabel{cgetkdc}
\insnmipslabel{csetkdc}
\subsubsection*{Get/Set Kernel Data Capability}

\begin{asmcode}
# The following are equivalent:
  CGetKDC $c1
  CReadHWR $30
\end{asmcode}

\begin{asmcode}
# The following are equivalent:
  CSetKDC $c1
  CWriteHWR $30
\end{asmcode}

\subsection{GGetKCC, CSetKCC}
\insnmipslabel{cgetkcc}
\insnmipslabel{csetkcc}
\subsubsection*{Get/Set Kernel Code Capability}

\begin{asmcode}
# The following are equivalent:
  CGetKCC $c1
  CReadHWR $29
\end{asmcode}

\begin{asmcode}
# The following are equivalent:
  CSetKCC $c1
  CWriteHWR $29
\end{asmcode}

\subsection{CAssertInBounds}
\insnmipslabel{cassertinbounds}
\subsubsection*{Assert that a capability in bounds}
This pseudo operation can be used to assert that a capability grants access to a given number of bytes (if the size argument is omitted, one byte is assumed).
This instruction only checks the bounds of the capability and ignores permissions. Therefore, an access might still be prohibited even if \insnnoref{CAssertInBounds} did not raise an exception.
\begin{asmcode}
# The following are equivalent (check that at least one byte is accessible):
  CAssertInBounds $c1
  CSetBoundsImm $cnull, $c1, 1
\end{asmcode}

\begin{asmcode}
# The following are equivalent (check that at least 10 bytes are accessible):
  CAssertInBounds $c1, 10
  CSetBoundsImm $cnull, $c1, 10
\end{asmcode}


\subsection{Capability Loads and Stores of Floating-Point Values}

The current revision of the CHERI ISA does not have CHERI-MIPS instructions for loading floating point values directly via capabilities.
MIPS does provide instructions for moving values between integer and floating point registers, so
a load or store of a floating point value via a capability
can be implemented in two instructions.

Four pseudo-instructions are defined to implement these patterns.
These are \asm{clwc1} and \asm{cldc1} for loading 32-bit and 64-bit floating point values, and \asm{cswc1} and \asm{csdc1} as the equivalent store operations.
The load operations expand as follows:
\begin{asmcode}
  cldc1    $f7, $zero, 0($c2)
  # Expands to:
  cld    $1, $zero, 0($c2)
  dmtc1    $1, $f7
\end{asmcode}

Note that integer register \regname{1} (\regname{at}) is used;
this pseudo-op is unavailable if the \asm{noat} directive is used.
The 32-bit variant (\asm{clwc1}) has a similar expansion, using \asm{clwu} and \asm{mtc1}.

The store operations are similar:

\begin{asmcode}
  csdc1    $f7, $zero, 0($c2)
  # Expands to:
  dmfc1    $1, $f7
  csd    $1, $zero, 0($c2)
\end{asmcode}

The specified floating point value is moved from the floating point register to \regname{at} and then stored using the correct-sized capability instruction.

\mrnote{If the floating point loads and stores were real instructions rather than
pseudo-ops, we would have to deal with the problem of what happens if a
capability exception or TLB exception is raised in the floating point pipeline.
The floating point and integer pipelines would need to be synchronized so that
the exception occurs at the correct point in program order.}
