\clearpage
\phantomsection
\addcontentsline{toc}{subsection}{CBEZ / CBNZ}
\insnmipslabel{cbez}
\subsection*{CBEZ / CBNZ: Branch if Capability is / is Not NULL}

\subsubsection*{Format}

CBEZ cb, offset

\begin{center}
\begin{bytefield}{32}
\bitheader[endianness=big]{0,15,16,20,21,25,26,31}\\
\bitbox{6}{0x12}
\bitbox{5}{0x11}
\bitbox{5}{cb}
\bitbox{16}{offset}
\end{bytefield}
\end{center}
\phantomsection
\insnmipslabel{cbnz}
CBNZ cb, offset

\begin{center}
\begin{bytefield}{32}
\bitheader[endianness=big]{0,15,16,20,21,25,26,31}\\
\bitbox{6}{0x12}
\bitbox{5}{0x12}
\bitbox{5}{cb}
\bitbox{16}{offset}
\end{bytefield}
\end{center}

\subsubsection*{Description}

Sets the \PC{} to \PC{} $+$ 4*{\em offset} $+$ 4, where {\em offset} is sign
extended, depending on whether \emph{cb} is equal to the NULL capability.

The instruction following the branch, in the delay slot, is executed before
branching.

\subsubsection*{Semantics}

\sailMIPScode{CBZ}

\subsubsection*{Notes}

\begin{itemize}
\item
In the above Sail code {\tt notzero} is false for \insnmipsref{CBEZ} and true
for \insnmipsref{CBNZ} thus inverting the sense of the
comparison (via exclusive-or) for the latter.
\item
Like all MIPS branch instructions, \insnmipsref{CBEZ} and
\insnmipsref{CBNZ} have a branch delay slot.  The instruction after
it will always be executed, regardless of whether the branch is taken
or not.
\item
This instruction is intended to resemble the conditional branch instructions
from the MIPS ISA. In particular, the shift left of the offset by 2 bits and
adding 4 is the same as MIPS conditional branches.
\item
Contrary to previous versions of the CHERI architecture the bounds check on
\PCC{} is performed during execution of the branch so an out-of-bounds target
will result in an exception. In the Sail code this check occurs in the
\lstinline{execute_branch} function.
\end{itemize}
