\clearpage
\phantomsection
\addcontentsline{toc}{subsection}{CClearTags}
\insnmipslabel{ccleartags}
\subsection*{CClearTags: Zero Multiple Capability Tags in Memory}

\subsubsection*{Format}

CClearTags rd, cb \\

\begin{center}
\begin{bytefield}{32}
\bitheader[endianness=big]{0,5,6,10,11,15,16,20,21,25,26,31}\\
\bitbox{6}{0x12}
\bitbox{5}{0x00}
\bitbox{5}{cb}
\bitbox{5}{0x18}
\bitbox{5}{0x1F}
\bitbox{5}{0x3F}
\end{bytefield}
\end{center}

\usesDDCinsteadofNULL{cb}

\subsubsection*{Description}

The \emph{tags} occupying the \emph{\insnmipsref{CLoadTags} stride} within
memory with base \emph{cb} are zeroed.  The result of this instruction must be
coherent with other processors, \emph{as if} the corresponding data memory
words had been stored, untagged, with their current contents.  The number of
tags zeroed is an implementation-defined constant but is equal to the number of
tags returned by \insnmipsref{CLoadTags}.

Capability register \emph{cb} must contain a capability that grants permission
to store data; capability storage is not required as the bits could just as
well be cleared by \insnmipsref[csbhwd]{CSD}.  The virtual address \emph{cb.\cbase{}}
$+$ \emph{cb.\coffset{}} must be suitably aligned, as with
\insnmipsref{CLoadTags}.  If any tag to be cleared corresponds to memory out of
bounds of \emph{cb}, a length violation is indicated.

\subsubsection*{Semantics}

\sailMIPScode{CClearTags}

\subsubsection*{Exceptions}

A coprocessor 2 exception is raised if:

\begin{itemize}
\item
\cchecktag{}
\item
\emph{cb.\ctag{}} is clear.
\item
\emph{cb} is sealed.
\item
\emph{cb}.\cperms.\emph{Permit\_Store} is not set.
\item
\emph{cb}.\cbase{} $+$ \emph{cb}.\coffset{} + $n$ $*$ \emph{capability\_size} $>$ \emph{cb}.\cbase{} $+$ \emph{cb}.\clength{},
		where $n$ is the number of capabilities to be cleared (\isail{caps_per_cacheline} in the Sail code).
\item
\emph{cb}.\cbase{} $+$ \emph{cb}.\coffset{} $<$ \emph{cb}.\cbase{}.
\end{itemize}

An address error during load (AdEL) exception is raised if:

\begin{itemize}
\item
The virtual address \emph{addr} is not $n$ $*$ \emph{capability\_size} aligned.
\end{itemize}
