\clearpage
\phantomsection
\addcontentsline{toc}{subsection}{CJR / CJALR}
\insnmipslabel{cjalr}
\subsection*{CJR / CJALR: Jump (and Link) Capability Register}

\subsubsection*{Format}

CJALR cb, cd

\begin{center}
\cheritwoop[header]{0xc}{cd}{cb}
\end{center}

\phantomsection
\insnmipslabel{cjr}

CJR cb

\begin{center}
\cherioneop[header]{0x3}{cb}
\end{center}

\note{rmn30}{Now that C0 is NULL we could encode this using CJALR with cd == 0 and save a whole opcode.}
\subsubsection*{Description}

The current \PCC{} (with an offset of the current \PC{} $+$ 8) is
optionally saved in \textit{cd}.
\PCC{} is then loaded from capability register \textit{cb} and \PC{}
is set from its offset.

\subsubsection*{Semantics}
\sailMIPScode{CJALR}

\subsubsection*{Exceptions}

A coprocessor 2 exception will be raised if:

\begin{itemize}
\item
\emph{cb.\ctag{}} is not set.
\item
\emph{cb} is sealed. (But see \cref{sec:arch-sentry}.)
\item
\textit{cb.\cperms.Permit\_Execute} is not set.
\item
\textit{cb.\coffset{}} + 4 is greater than \textit{cb.\clength{}}.
\end{itemize}

An address error exception will be raised if

\begin{itemize}
\item
\textit{cb.\cbase{}} + \textit{cb.\coffset{}} is not 4-byte word aligned.
\end{itemize}

\subsubsection*{Notes}

\begin{itemize}
\item
\insnmipsref{CJALR} has a branch delay slot.
\item
The change to \PCC{} does not take effect until the instruction in the branch
delay slot has been executed.
\item
\note{rmn30}{The following text is copied from cjr but was not present here before I merged cjr and cjalr. What is the importance of this? I think it is just to rule out offsets effectively less than zero but since we actually store the address (not offset) and compare against base in Sail I think it is redundant.}
\textit{cb.\cbase{}} and \textit{cb.\clength{}} are treated as unsigned integers,
and the result of the addition does not wrap around (i.e., an exception is
raised if \emph{cb.\cbase{}}+\emph{cb.\coffset{}} is greater than maxaddr).
\end{itemize}
