\clearpage
\phantomsection
\addcontentsline{toc}{subsection}{CLC}
\insnmipslabel{clc}
\label{sailMIPSzCLC} % for CLCNT
\subsection*{CLC: Load Capability via Capability}

\subsubsection*{Format}

CLC cd, rt, offset(cb) \\
CLCR cd, rt(cb) \\
CLCI cd, offset(cb)

\begin{center}
\begin{bytefield}{32}
\bitheader[endianness=big]{0,5,6,10,11,15,16,20,21,25,26,31}\\
\bitbox{6}{0x36}
\bitbox{5}{cd}
\bitbox{5}{cb}
\bitbox{5}{rt}
\bitbox{11}{offset}
\end{bytefield}
\end{center}

\usesDDCinsteadofNULL{cb}

\subsubsection*{Description}
Capability register \emph{cd} is loaded from the memory location specified by
\emph{cb.\cbase{}} $+$ \emph{cb.\coffset{}} $+$ \emph{rt} $+$ \emph{offset}.
Capability register
\emph{cb} must
contain a capability that grants permission to load capabilities.  The virtual
address \emph{cb.\cbase{}} $+$ \emph{cb.\coffset{}} $+$ \emph{rt} $+$
\emph{offset} must be \emph{capability\_size} aligned.

The bit in the tag memory corresponding to \emph{cb.\cbase{}} $+$
\emph{cb.\coffset{}} $+$ \emph{rt} $+$ \emph{offset} is
loaded into the tag bit associated with \emph{cd}.
%  The CBTU instruction can be used to check the value of this tag.

\subsubsection*{Semantics}
\sailMIPScode{CLC}

\subsubsection*{Exceptions}


A coprocessor 2 exception is raised if:

\begin{itemize}
\item
\cchecktag{}
\item
\emph{cb} is sealed.
\item
\emph{cb}.\cperms.\emph{Permit\_Load} is not set.
\item
\emph{addr} + \emph{capability\_size} $>$ \emph{cb}.\cbase{} $+$ \emph{cb}.\clength{}.
\item
\emph{addr} $<$ \emph{cb}.\cbase{}.
\end{itemize}


An address error during load (AdEL) exception is raised if:

\begin{itemize}
\item
The virtual address \emph{addr} is not \emph{capability\_size} aligned.
\end{itemize}

\subsubsection*{Notes}

\begin{itemize}
\item
This instruction reuses the opcode from the Load Doubleword to Coprocessor 2
(\insnnoref{LDC2}) instruction in the MIPS Specification.
\item
\emph{offset} is interpreted as a signed integer.
\item
The \insnmipsref[clbhwd]{CLCI} mnemonic is equivalent to \insnmipsref{CLC} with
\emph{cb} being the zero register (\$zero). The \insnmipsref[clbhwd]{CLCR} mnemonic
is equivalent to \insnmipsref{CLC} with \emph{offset} set to
zero.
\item
Although the \emph{capability\_size} can vary, the offset is always in
multiples of 16 bytes (128 bits).
\end{itemize}
