\clearpage
\phantomsection
\addcontentsline{toc}{subsection}{CLShC}
\insnmipslabel{clshc}
\insnmipslabel{clshci}
\insnmipslabel{clshcr}
\subsection*{CLShC: Load Short Capability via Capability}

\subsubsection*{Format}

CLShC cd, rt, offset(cb) \\
CLShCR cd, rt(cb) \\
CLShCI cd, offset(cb)

\usesDDCinsteadofNULL{cb}

\subsubsection*{Description}

Capability register \emph{cd} is loaded with the decoded form of the short
capability located at the memory location specified by
%
\emph{addr} $\stackrel{\Delta}{=}$ \emph{cb.\cbase{}} $+$
\emph{cb.\coffset{}} $+$ \emph{rt} $+$ 8 $*$ \emph{offset},
%
provided this address is suitably aligned for short capabilities.
%
Capability register \emph{cb} must contain a capability that grants
permission to load capabilities.

The tag bit associated with \emph{cd} is set if the tag bits associated with
\emph{addr} indicate that this address contain a short capability and
\emph{cb} bears Permit\_Load\_Capability.  See
\cref{sec:windowedshortcaps} for possible tag representations.

% \subsubsection*{Semantics}

\subsubsection*{Exceptions}

A coprocessor 2 exception is raised if:

\begin{itemize}
\item
\cchecktag{}
\item
\emph{cb} is sealed.
\item
\emph{cb}.\cperms.\emph{Permit\_Load} is not set.
\item
\emph{addr} + \emph{short\_capability\_size} $>$ \emph{cb}.\cbase{} $+$ \emph{cb}.\clength{}.
\item
\emph{addr} $<$ \emph{cb}.\cbase{}.
\end{itemize}

An address error during load (AdEL) exception is raised if:

\begin{itemize}
\item
The virtual address \emph{addr} is not \emph{capability\_size} aligned.
\end{itemize}

\subsubsection*{Notes}

\begin{itemize}
\item
\emph{offset} is interpreted as a signed integer.
\item
The \insnmipsref{CLShCI} mnemonic is equivalent to \insnmipsref{CLShC} with
\emph{cb} being the zero register (\$zero). The \insnmipsref{CLShCR} mnemonic
is equivalent to \insnmipsref{CLShC} with \emph{offset} set to
zero.
\item
Although the \emph{short\_capability\_size} may vary, the offset is always in
multiples of 8 bytes (64 bits).
\item
See \cref{sec:windowedshortcaps} for details of short capabilities.  In
particular, note that several permissions are not transported in short
capabilities.
\end{itemize}
