function #\hyperref[sailRISCVzlegalizzezymstatus]{legalize\_mstatus}#(o : Mstatus, v : xlenbits) -> Mstatus = {
  /*
   * Populate all defined fields using the bits of v, stripping anything
   * that does not have a matching bitfield entry. All bits above 32 are handled
   * explicitly later.
   */
  let m : Mstatus = #\hyperref[sailRISCVzMkzyMstatus]{Mk\_Mstatus}#(#\hyperref[sailRISCVzEXTZ]{EXTZ}#(v[22 .. 11] @ 0b00 @ v[8 .. 7] @ 0b0 @ v[5 .. 3] @ 0b0 @ v[1 .. 0]));

  /* We don't have any extension context yet. */
  let m = #\hyperref[sailRISCVzupdatezyXS]{update\_XS}#(m, #\hyperref[sailRISCVzextStatuszytozybits]{extStatus\_to\_bits}#(Off));

  /* FS is WARL, and making FS writable can support the M-mode emulation of an FPU
   * to support code running in S/U-modes.  Spike does this, and for now, we match it,
   * but only if Zfinx isn't enabled.
   * FIXME: This should be made a platform parameter.
   */
  let m = if #\hyperref[sailRISCVzsyszyenablezyzzfinx]{sys\_enable\_zfinx}#() then #\hyperref[sailRISCVzupdatezyFS]{update\_FS}#(m, #\hyperref[sailRISCVzextStatuszytozybits]{extStatus\_to\_bits}#(Off)) else m;
  let dirty = #\hyperref[sailRISCVzextStatuszyofzybits]{extStatus\_of\_bits}#(m.#\hyperref[sailRISCVzFS]{FS}#()) == Dirty | #\hyperref[sailRISCVzextStatuszyofzybits]{extStatus\_of\_bits}#(m.#\hyperref[sailRISCVzXS]{XS}#()) == Dirty;
  let m = #\hyperref[sailRISCVzupdatezySD]{update\_SD}#(m, #\hyperref[sailRISCVzboolzytozybits]{bool\_to\_bits}#(dirty));

  /* We don't support dynamic changes to SXL and UXL. */
  let m = #\hyperref[sailRISCVzsetzymstatuszySXL]{set\_mstatus\_SXL}#(m, #\hyperref[sailRISCVzgetzymstatuszySXL]{get\_mstatus\_SXL}#(o));
  let m = #\hyperref[sailRISCVzsetzymstatuszyUXL]{set\_mstatus\_UXL}#(m, #\hyperref[sailRISCVzgetzymstatuszyUXL]{get\_mstatus\_UXL}#(o));

  /* We don't currently support changing MBE and SBE. */
  let m = if sizeof(xlen) == 64 then {
             #\hyperref[sailRISCVzMkzyMstatus]{Mk\_Mstatus}#([m.#\hyperref[sailRISCVzbits]{bits}#() with 37 .. 36 = 0b00])
          } else m;

  /* Hardwired to zero in the absence of 'U' or 'N'. */
  let m = if #\hyperref[sailRISCVznot]{not}#(#\hyperref[sailRISCVzhaveNExt]{haveNExt}#()) then {
             let m = #\hyperref[sailRISCVzupdatezyUPIE]{update\_UPIE}#(m, 0b0);
             let m = #\hyperref[sailRISCVzupdatezyUIE]{update\_UIE}#(m, 0b0);
             m
          } else m;

  if #\hyperref[sailRISCVznot]{not}#(#\hyperref[sailRISCVzhaveUsrMode]{haveUsrMode}#()) then {
    let m = #\hyperref[sailRISCVzupdatezyMPRV]{update\_MPRV}#(m, 0b0);
    m
  } else m
}
